\texttt{gwsnr}: A Python Package for Efficient SNR Calculation of GWs

### 2024-11-09 arxiv/paper.tex
mac

Gravitational waves (GWs), ripples in spacetime predicted by Einstein's
theory of General Relativity, have revolutionized astrophysics since
their first detection in 2015 (Abbott, B. P. et al. (2016a)). These
waves are emitted by cataclysmic events like the merging of binary black
holes (BBHs), binary neutron stars (BNSs) and BH-NS pairs, providing a
unique window into the cosmos. A critical aspect of GW analysis is the
Signal-to-Noise Ratio (SNR). SNR quantifies the strength of a GW signal
relative to the background noise in a detector, like LIGO (The LIGO
Scientific Collaboration et al. (2015)) or Virgo (Acernese et al.
(2014)). This ratio is pivotal in confirming the detection of GWs and
extracting astrophysical information from them (Abbott, B. P. et al.
(2016b)). However, specific scenarios in GW research, particularly in
simulations of detectable GW events (Abbott et al. (2016)) and in
hierarchical Bayesian analysis (Thrane and Talbot (2019)) where
selection effects are considered, demand extensive and efficient
computation of SNR. This requirement presents a significant challenge,
as conventional computational approaches, such as noise-weighted inner
product, are typically time-consuming and impractical for such
specialized and large-scale analyses (Taylor and Gerosa (2018), Gerosa
et al. (2020)).

@@

The \emph{\texttt{gwsnr}} Python package addresses the need for
efficient SNR computation in GW research. It innovatively streamlines
SNR calculations, enhancing accuracy and efficiency with several
advanced techniques. Firstly, it utilizes an innovative interpolation
method, employing a partial-scaling approach for accurately
interpolating the SNR of GWs from spin-less binary systems wrt to their
mass parameters. Secondly, the package features a noise-weighted inner
product method, similar to that in the \emph{\texttt{bilby}} package
(Ashton, Gregory et al. (2022)), but enhanced with multiprocessing
capabilities. This integration allows for the parallel processing of
complex calculations, thereby expediting the SNR computation. Thirdly, a
trained Artificial Neural Network (ANN) model is incorporated for rapid
`probability of detection' (Pdet) estimation for BBH systems with spin
precession. Lastly, \emph{\texttt{gwsnr}} leverages the
\emph{\texttt{numba}}'s Just-In-Time (njit) compiler (Lam, Pitrou, and
Seibert (2022)), which optimizes performance by compiling Python code
into machine code at runtime, drastically reducing execution times.
Beyond these technical merits, \emph{\texttt{gwsnr}} stands out for its
user-friendly features and seamless integration with other related
software packages, making it not just a powerful tool but also an
accessible one for researchers. These attributes position
\emph{\texttt{gwsnr}} as an invaluable asset in GW data analysis,
particularly in simulations of detectable compact binary mergers,
determining rates of GW events, ascertain to both lensed and un-lensed
events (Phurailatpam et al. (2024), Ng et al. (2024)), and in the
analysis of selection effects within hierarchical Bayesian frameworks
(Thrane and Talbot (2019)). The package thus represents a significant
step forward in aiding GW research, enabling more precise and efficient
exploration of the universe through GW observations and simulations.
Additionally, \emph{\texttt{gwsnr}} is instrumental in the
\emph{\texttt{ler}} (Phurailatpam et al. (2024)) package for calculating
detectable rates of both lensed and unlensed GW events, showcasing its
utility in advanced GW studies.

@@

  \begin{itemize}
  \item
    For a simple inspiral waveform, the optimal SNR is given by,
    \[\rho = F(D_l,\mathcal{M},\iota,\psi,\alpha, \delta) \sqrt{ 4\int_0^{f_{ISCO}} \frac{f^{-7/3}}{S_n(f)}df }\]
  \item
    \(F\) is defined as a function of luminosity distance (\(D_l\)),
    chirp mass (\(\mathcal{M}\)), inclination angle (\(\iota\)),
    polarization angles (\(\psi\)), right ascension (\(\alpha\)), and
    declination (\(\delta\)). \(f\) is the frequency, \(f_{ISCO}\) is
    the last stable orbit frequency and \(S_n(f)\) is the detector's
    noise curve or power spectral density (psd).
  \item
    Then, partial scaled SNR:
    \(\rho_{1/2} = \sqrt{ 4\int_0^\infty \frac{f^{-7/3}}{S_n(f)}df } \approx \sqrt{ 4\int_0^{f_{ISCO}} \frac{f^{-7/3}}{S_n(f)}df }\)
  \item
    For an spinless frequency-domain IMR waveform with optimal SNR equal
    to \(\rho\):
    \(\rho_{1/2} = \rho\,/\, F(D_l,\mathcal{M},\iota,\psi,\alpha, \delta, \psi)\)
  \item
    \(\rho_{1/2}\) is considered a function of \(M_{tot}\) and \(q\).
  \end{itemize}
\end{itemize}

@@

@@

@@

>>>>

Gravitational waves (GWs), ripples in spacetime predicted by Einstein's
theory of General Relativity, have revolutionized astrophysics since
their first detection in 2015 (Abbott, B. P. et al. (2016a)). These
waves are emitted by cataclysmic events like the merging of binary black
holes (BBHs), binary neutron stars (BNSs) and BH-NS pairs, providing a
unique window into the cosmos. A critical aspect of GW analysis is the
Signal-to-Noise Ratio (SNR). SNR quantifies the strength of a GW signal
relative to the background noise in a detector, like LIGO (The LIGO
Scientific Collaboration et al. (2015), Abbott et al. (2020), Buikema et
al. (2020)), Virgo (F. Acernese et al. (2014), F. Acernese et al.
(2019)) or KAGRA (Akutsu et al. (2020), Aso et al. (2013)). This ratio
is pivotal in confirming the detection of GWs and extracting
astrophysical information from them (Abbott, B. P. et al. (2016b)).
However, specific scenarios in GW research, particularly in simulations
of detectable GW events (Abbott et al. (2016)) and in hierarchical
Bayesian analysis (Thrane and Talbot (2019)) where selection effects are
considered, demand extensive and efficient computation of SNR. This
requirement presents a significant challenge, as conventional
computational approaches, such as noise-weighted inner product, are
typically time-consuming and impractical for such specialized and
large-scale analyses (Taylor and Gerosa (2018), Gerosa et al. (2020)).

@@

The \emph{\texttt{gwsnr}} Python package addresses the need for
efficient SNR computation in GW research. It innovatively streamlines
SNR calculations, enhancing accuracy and efficiency with several
advanced techniques. Firstly, it utilizes an innovative interpolation
method, employing a partial-scaling approach for accurately
interpolating the SNR of GWs from spin-less binary systems wrt to their
mass parameters. Secondly, the package features a noise-weighted inner
product method, similar to that in the \emph{\texttt{bilby}} package
(Ashton, Gregory et al. (2022)), but enhanced with multiprocessing
capabilities. This integration allows for the parallel processing of
complex calculations, thereby expediting the SNR computation. Thirdly, a
trained Artificial Neural Network (ANN) model is incorporated for rapid
`probability of detection' (Pdet) estimation for BBH systems with spin
precession. Lastly, \emph{\texttt{gwsnr}} leverages the
\emph{\texttt{numba}}'s Just-In-Time (njit) compiler (Lam, Pitrou, and
Seibert (2022)), which optimizes performance by compiling Python code
into machine code at runtime, drastically reducing execution times.
Beyond these technical merits, \emph{\texttt{gwsnr}} stands out for its
user-friendly features and seamless integration with other related
software packages, making it not just a powerful tool but also an
accessible one for researchers. These attributes position
\emph{\texttt{gwsnr}} as an invaluable asset in GW data analysis,
particularly in simulations of detectable compact binary mergers,
determining rates of GW events, related to both lensed and un-lensed
events (Phurailatpam et al. (2024), Ng et al. (2024)), and in the
analysis of selection effects within hierarchical Bayesian frameworks
(Thrane and Talbot (2019)). The package thus represents a significant
step forward in aiding GW research, enabling more precise and efficient
exploration of the universe through GW observations and simulations.
Additionally, \emph{\texttt{gwsnr}} is instrumental in the
\emph{\texttt{ler}} (Phurailatpam et al. (2024)) package for calculating
detectable rates of both lensed and unlensed GW events, showcasing its
utility in advanced GW studies.

@@

  \begin{itemize}
  \item
    For a simple inspiral waveform, the optimal SNR is given by,
    \[\rho = F(D_l,\mathcal{M},\iota,\psi,\alpha, \delta) \sqrt{ 4\int_0^{f_{ISCO}} \frac{f^{-7/3}}{S_n(f)}df }\]
  \item
    \(F\) is defined as a function of luminosity distance (\(D_l\)),
    chirp mass (\(\mathcal{M}\)), inclination angle (\(\iota\)),
    polarization angles (\(\psi\)), right ascension (\(\alpha\)), and
    declination (\(\delta\)), refer to Eqn(D1) of Allen et al. (2012).
    \(f\) is the frequency, \(f_{ISCO}\) is the last stable orbit
    frequency and \(S_n(f)\) is the detector's noise curve or power
    spectral density (psd).
  \item
    Then, partial scaled SNR:
    \(\rho_{1/2} = \sqrt{ 4\int_0^\infty \frac{f^{-7/3}}{S_n(f)}df } \approx \sqrt{ 4\int_0^{f_{ISCO}} \frac{f^{-7/3}}{S_n(f)}df }\)
  \item
    For an spinless frequency-domain IMR waveform with optimal SNR equal
    to \(\rho\):
    \(\rho_{1/2} = \rho\,/\, F(D_l,\mathcal{M},\iota,\psi,\alpha, \delta)\)
  \item
    \(\rho_{1/2}\) is considered a function of \(M_{tot}\) and \(q\).
  \end{itemize}
\end{itemize}

@@

\bibitem[\citeproctext]{ref-Abbott2020}
Abbott, B. P., R. Abbott, T. D. Abbott, S. Abraham, F. Acernese, K.
Ackley, C. Adams, et al. 2020. {``Prospects for Observing and Localizing
Gravitational-Wave Transients with Advanced LIGO, Advanced Virgo and
KAGRA.''} \emph{Living Reviews in Relativity} 23 (1).
\url{https://doi.org/10.1007/s41114-020-00026-9}.

\bibitem[\citeproctext]{ref-virgo:2019}
Acernese, F., M. Agathos, L. Aiello, A. Allocca, A. Amato, S. Ansoldi,
S. Antier, et al. 2019. {``Increasing the Astrophysical Reach of the
Advanced Virgo Detector via the Application of Squeezed Vacuum States of
Light.''} \emph{Phys. Rev. Lett.} 123 (December): 231108.
\url{https://doi.org/10.1103/PhysRevLett.123.231108}.

@@

\bibitem[\citeproctext]{ref-akutsu2020overviewkagradetectordesign}
Akutsu, T., M. Ando, K. Arai, Y. Arai, S. Araki, A. Araya, N. Aritomi,
et al. 2020. {``Overview of KAGRA: Detector Design and Construction
History.''} \url{https://arxiv.org/abs/2005.05574}.

@@

\bibitem[\citeproctext]{ref-PhysRevD:88:043007}
Aso, Yoichi, Yuta Michimura, Kentaro Somiya, Masaki Ando, Osamu
Miyakawa, Takanori Sekiguchi, Daisuke Tatsumi, and Hiroaki Yamamoto.
2013. {``Interferometer Design of the KAGRA Gravitational Wave
Detector.''} \emph{Phys. Rev. D} 88 (August): 043007.
\url{https://doi.org/10.1103/PhysRevD.88.043007}.

\bibitem[\citeproctext]{ref-PhysRevD.102.062003}
Buikema, A., C. Cahillane, G. L. Mansell, C. D. Blair, R. Abbott, C.
Adams, R. X. Adhikari, et al. 2020. {``Sensitivity and Performance of
the Advanced LIGO Detectors in the Third Observing Run.''} \emph{Phys.
Rev. D} 102 (September): 062003.
\url{https://doi.org/10.1103/PhysRevD.102.062003}.
